---
title: "covidDraftV2"
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cols4all)
library(here)
```

### Question 1

#### 1.1

**Download the dataset: download Genomes_per_week_in_England.csv from Canvas. This includes weekly counts of virus samples per lineage over time across England collected as part of Sanger Institute COG-UK.**

```{r downloadDataset}

# Import COG-UK daily genomic sequence data
COGdata <- read.csv(here("data", "Genomes_per_week_in_England.csv"))

# Check the input format and display the first few rows of the data
head(COGdata)
```

#### 1.2

**Classify major lineages: identify the following variants as major lineages: B.1.1.7 (Alpha), B.1.617.2 (Delta), BA.1, BA.2, BA.2.75, BA.4, BA.5, BA.5.3 (BQ.1), and XBB. Group all other lineages into a single category labelled as Other.**

```{r setMajorLineages}

# Specify the major lineages
major_lineages <- c("B.1.1.7", "B.1.617.2", "BA.1", "BA.2", "BA.2.75", "BA.4", "BA.5", "BA.5.3", "XBB")

# Cleaning the data
COGclean <- COGdata %>%
  # Ensure collection_date is in Date format
  mutate(date = as.Date(date)) %>%
  # Grouping all non-major lineages as 'other' 
  mutate(lineage = ifelse(lineage %in% major_lineages, lineage, "Other")) %>%
  # Rename columns for clarity
  {colnames(.) <- c("collection_date", "major_lineage", "lineage_count"); .}

# Check the format of the columns after cleaning the data and displaying the first few rows of the data

head(COGclean)
```

#### 1.3

**Visualise the data: generate a stacked area plot showing the total counts of each major lineage over time. Generate another stacked area plot showing the frequencies (proportions) of each major lineage over time.**

```{r frequencyData}

# Calculate total counts per date
total_COG_counts <- aggregate(COGclean$lineage_count, by = list(collection_date = COGclean$collection_date), FUN = sum)
colnames(total_COG_counts) <- c("collection_date", "total_count")

# Merge total counts back into the lineage summary
COGclean <- merge(COGclean, total_COG_counts, by = "collection_date")

# Calculate frequencies
COGclean$lineage_frequency <- COGclean$lineage_count / COGclean$total_count

# Display the first few rows of the new data frame
head(COGclean)
```

```{r 10dayBins}
# Aggregate lineage frequencies into 10-day bins
# Converts the Date values into numeric format, where each date is represented as the number of days since 1970-01-01 (the Unix epoch)
COGclean$collection_date_bin <- as.Date(
  floor(as.numeric(as.Date(COGclean$collection_date)) / 7) * 7, origin = "1970-01-01"
)

# Aggregate lineage counts for each 10-day bin
COG_summary_binned <- aggregate(
  lineage_count ~ collection_date_bin + major_lineage,
  data = COGclean,
  FUN = sum
)

# Calculate total counts within each bin
total_COG_counts <- aggregate(
  lineage_count ~ collection_date_bin,
  data = COG_summary_binned,
  FUN = sum
)
colnames(total_COG_counts) <- c("collection_date_bin", "total_count")  # Rename for clarity

# Merge total counts back into the binned data
COG_summary_binned <- merge(COG_summary_binned, total_COG_counts, by = "collection_date_bin")

# Recalculate frequencies
COG_summary_binned$lineage_frequency <- COG_summary_binned$lineage_count / COG_summary_binned$total_count

# Preview the binned data
head(COG_summary_binned)

```

```{r stackedAreaPlotCounts}

# Set a colour blind friendly colour palette
palette <- c4a("safe",10)

# Make stacked area plot for total counts of each major lineage
ggplot(COGclean, aes(x = collection_date, y = lineage_count, fill = major_lineage)) +
    geom_area(position = "fill") +  # Create a stacked area plot
    scale_fill_manual(values = palette) +  # Apply the custom palette
    labs(
      title = "Total counts of major lineages over time",
      x = "Collection date",
      y = "Total count",
      fill = "Major lineage"
    ) +
    theme_minimal()

##this doesn't seem quite right?
```

```{r stackedAreaPlotFrequencies}

# Make stacked area plot for frequencies of each major lineage
ggplot(COG_summary_binned, aes(x = collection_date_bin, y = lineage_frequency, fill = major_lineage)) +
    geom_area(position = "fill") +  # Create a stacked area plot
    scale_fill_manual(values = palette) +  # Apply the custom palette
    labs(
      title = "Frequency of major lineages over time",
      x = "Collection date",
      y = "Proportion",
      fill = "Major lineage"
    ) +
    theme_minimal()


```


### Question 2

####2.1
**Visualise the COG-UK and ONS-CIS data for BA.2: plot the frequency trajectory for the BA.2 variant using both the Sanger dataset (weekly counts) and the ONS-CIS dataset (10-day bin counts from the practical).**

```{r ONSdataSetup}

# Import ONS-CIS daily genomic sequence data
ONSdata <- read.csv("https://raw.githubusercontent.com/mg878/variant_fitness_practical/main/lineage_data.csv")

# Ensure collection_date is in Date format
ONSdata$collection_date <- as.Date(ONSdata$collection_date)

# Create new data frame 
ONSsummary <- aggregate(
  ONSdata$major_lineage,
  by = list(collection_date = ONSdata$collection_date, major_lineage = ONSdata$major_lineage),
  FUN = length
)

# Rename columns for clarity
colnames(ONSsummary) <- c("collection_date", "major_lineage", "lineage_count")

# Calculate total counts per date
total_ONS_counts <- aggregate(ONSsummary$lineage_count, by = list(collection_date = ONSsummary$collection_date), FUN = sum)
colnames(total_ONS_counts) <- c("collection_date", "total_count")

# Merge total counts back into the lineage summary
ONSsummary <- merge(ONSsummary, total_ONS_counts, by = "collection_date")

# Calculate frequencies
ONSsummary$lineage_frequency <- ONSsummary$lineage_count / ONSsummary$total_count

# Display the first few rows of the new data frame
head(ONSsummary)
```

```{r ONS10dayBins}
# Aggregate lineage frequencies into 10-day bins
# Converts the Date values into numeric format, where each date is represented as the number of days since 1970-01-01 (the Unix epoch)
ONSsummary$collection_date_bin <- as.Date(
  floor(as.numeric(as.Date(ONSsummary$collection_date)) / 10) * 10, origin = "1970-01-01"
)

# Aggregate lineage counts for each 10-day bin
ONS_summary_binned <- aggregate(
  lineage_count ~ collection_date_bin + major_lineage,
  data = ONSsummary,
  FUN = sum
)

# Calculate total counts within each bin
total_ONS_counts <- aggregate(
  lineage_count ~ collection_date_bin,
  data = ONS_summary_binned,
  FUN = sum
)

# Rename columns for clarity
colnames(total_ONS_counts) <- c("collection_date_bin", "total_count") 

# Merge total counts back into the binned data
ONS_summary_binned <- merge(ONS_summary_binned, total_ONS_counts, by = "collection_date_bin")

# Recalculate frequencies
ONS_summary_binned$lineage_frequency <- ONS_summary_binned$lineage_count / ONS_summary_binned$total_count

# Preview the binned data
head(ONS_summary_binned)

```

```{r BA.2frequencyTrajectory}

# Filter COG data for BA.2
COG_daily_trajectories <- subset(
  COG_summary_binned,
  major_lineage %in% c("BA.2")
)

# Filter ONS data for BA.2
ONS_daily_trajectories <-subset(
  ONS_summary_binned,
  major_lineage %in% c("BA.2")
)

#attempting to sort out the second line
ggplot(COG_daily_trajectories, aes(x = collection_date_bin, y = lineage_frequency)) +
  geom_line(data = COG_daily_trajectories, linewidth = 1, color = "blue" ) +  # Connect the points with lines
  geom_line(data = ONS_daily_trajectories, linewidth = 1, color = "red")
#   geom_point(size = 2, alpha = 0.7) +  # Add points for daily frequencies
#   scale_color_manual(
#     values = c("BA.2" = "blue"),
#     labels = c("BA.2 (Alpha)")
#   ) +
#   geom_line(data = ONS_daily_trajectories)
#   labs(
#     title = "Daily frequency trajectories of BA.2",
#     x = "Collection date",
#     y = "Proportion",
#     color = "Major lineage"
#   ) +
#   theme_minimal()

#i know this one works but the lines aren't quite right
ggplot(COG_daily_trajectories, aes(x = collection_date_bin, y = lineage_frequency, color = major_lineage, group = major_lineage)) +
  geom_line(linewidth = 1) +  # Connect the points with lines
  geom_point(size = 2, alpha = 0.7) +  # Add points for daily frequencies
  scale_color_manual(
    values = c("BA.2" = "blue"),
    labels = c("BA.2 (Alpha)")
  ) +
  geom_line(data = ONS_daily_trajectories)
  labs(
    title = "Daily frequency trajectories of BA.2",
    x = "Collection date",
    y = "Proportion",
    color = "Major lineage"
  ) +
  theme_minimal()
```

