---
title: "covidDraftV2"
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, cols4all)
```

### Question 1

#### 1.1

**Download the dataset: download Genomes_per_week_in_England.csv from Canvas. This includes weekly counts of virus samples per lineage over time across England collected as part of Sanger Institute COG-UK.**

```{r downloadDataset}

# Import COG-UK daily genomic sequence data
COGdata <- read.csv("Genomes_per_week_in_England.csv")

# Check the input format and display the first few rows of the data
head(COGdata)
```

#### 1.2

**Classify major lineages: identify the following variants as major lineages: B.1.1.7 (Alpha), B.1.617.2 (Delta), BA.1, BA.2, BA.2.75, BA.4, BA.5, BA.5.3 (BQ.1), and XBB. Group all other lineages into a single category labelled as Other.**

```{r setMajorLineages}

# Specify the major lineages
major_lineages <- c("B.1.1.7", "B.1.617.2", "BA.1", "BA.2", "BA.2.75", "BA.4", "BA.5", "BA.5.3", "XBB")

# Cleaning the data
COGclean <- COGdata %>%
  # Ensure collection_date is in Date format
  mutate(date = as.Date(date)) %>%
  # Grouping all non-major lineages as 'other' 
  mutate(lineage = ifelse(lineage %in% major_lineages, lineage, "Other")) %>%
  # Rename columns for clarity
  {colnames(.) <- c("collection_date", "major_lineage", "lineage_count"); .}

# Check the format of the columns after cleaning the data and displaying the first few rows of the data

head(COGclean)
```

#### 1.3

**Visualise the data: generate a stacked area plot showing the total counts of each major lineage over time. Generate another stacked area plot showing the frequencies (proportions) of each major lineage over time.**

```{r frequencyData}

# Calculate total counts per date
total_COG_counts <- aggregate(COGclean$lineage_count, by = list(collection_date = COGclean$collection_date), FUN = sum)
colnames(total_COG_counts) <- c("collection_date", "total_count")

# Merge total counts back into the lineage summary
COGclean <- merge(COGclean, total_COG_counts, by = "collection_date")

# Calculate frequencies
COGclean$lineage_frequency <- COGclean$lineage_count / COGclean$total_count

# Display the first few rows of the new data frame
head(COGclean)
```

```{r 10dayBins}
# Aggregate lineage frequencies into 10-day bins
# Converts the Date values into numeric format, where each date is represented as the number of days since 1970-01-01 (the Unix epoch)
COGclean$collection_date_bin <- as.Date(
  floor(as.numeric(as.Date(COGclean$collection_date)) / 10) * 10, origin = "1970-01-01"
)

# Aggregate lineage counts for each 10-day bin
COG_summary_binned <- aggregate(
  lineage_count ~ collection_date_bin + major_lineage,
  data = COGclean,
  FUN = sum
)

# Calculate total counts within each bin
total_counts <- aggregate(
  lineage_count ~ collection_date_bin,
  data = COG_summary_binned,
  FUN = sum
)
colnames(total_counts) <- c("collection_date_bin", "total_count")  # Rename for clarity

# Merge total counts back into the binned data
COG_summary_binned <- merge(COG_summary_binned, total_counts, by = "collection_date_bin")

# Recalculate frequencies
COG_summary_binned$lineage_frequency <- COG_summary_binned$lineage_count / COG_summary_binned$total_count

# Preview the binned data
head(COG_summary_binned)

```

```{r stackedAreaPlotCounts}

# Set a colour blind friendly colour palette
palette <- c4a("safe",10)

# Make stacked area plot for total counts of each major lineage
ggplot(COGclean, aes(x = collection_date, y = lineage_count, fill = major_lineage)) +
    geom_area(position = "fill") +  # Create a stacked area plot
    scale_fill_manual(values = palette) +  # Apply the custom palette
    labs(
      title = "Total counts of major lineages over time",
      x = "Collection date",
      y = "Total count",
      fill = "Major lineage"
    ) +
    theme_minimal()

##this doesn't seem quite right?
```

```{r stackedAreaPlotFrequencies}

# Make stacked area plot for frequencies of each major lineage
ggplot(COG_summary_binned, aes(x = collection_date_bin, y = lineage_frequency, fill = major_lineage)) +
    geom_area(position = "fill") +  # Create a stacked area plot
    scale_fill_manual(values = palette) +  # Apply the custom palette
    labs(
      title = "Frequency of major lineages over time",
      x = "Collection date",
      y = "Proportion",
      fill = "Major lineage"
    ) +
    theme_minimal()


```
